Non-word repetition in Yélî Dnye

### 2020-12-21 manuscript.tex
add item complexity disc subsection draft

Next, we investigated patterns of deletion and substitution. Deletions
were relatively rare: there were only 7 instances of deleted vowels
(\textasciitilde{}0.23\% of all vowel targets), and 3 instances of
deleted consonants (\textasciitilde{}0.12\% of all consonant targets).
Substitutions were somewhat more frequent, but still relatively rare.
Nasal vowels were produced as oral vowels and vice versa (34 oral
targets as nasal (\textasciitilde{}0.99\% of targets) and 30 nasal
targets produced as oral (\textasciitilde{}2.37\% of targets). Both oral
and nasal vowels were also occasionally subsituted by vowels of a
different quality that preserved the target oral/nature airflow (110
oral vowels (\textasciitilde{}1.86\% of targets and 13 nasal vowels
(\textasciitilde{}2.06\% of targets). Among consonants, complex
consonants were occasionally substituted by simpler ones (specifically
{[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and
{[}lβʲ{]} as {[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]},
{[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and
{[}tʃ{]}), 115 times, in approximately 1.96\% of all complex consonant
targets. Interestingly, simple consonants were sometimes also
substituted by complex ones; this happened 2 times, in approximately
1.96\% of all simple consonant targets. Finally, simple consonants were
mispronounced as other simple consonants (68 instances,
\textasciitilde{}0.38\% of all simple consonant targets) and complex
consonants as other complex consonants (33 instances,
\textasciitilde{}1.02\% of all complex consonant targets). Note that the
substitutions included phones that are not native to Yélî Dnye but do
occur in English (e.g., {[}tʃ{]}); as these data come from careful
transcriptions by a native Yélî Dnye speaker who is very fluent in
English, we take these segmental substitutions faithfully as an
indication that several of our participants have mastered production of
some English phones, possibly produced within whole English word forms
(see below).

@@

\begin{itemize}
\item
  What is the overall repetition scores (whole word, phoneme based,
  distance)?
\item
  How does this change as a function of item complexity (number of
  syllables, sound complexity)?
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Prediction based on previous work: Children have higher scores for
  shorter items
\item
  Prediction for Yélî made before seeing the data: The length
  distribution in Yélî words is more balanced than that in English, and
  thus the score decline for poly- versus mono-syllables may be less
  pronounced than that for English.
\item
  it turns out we were right! the scores for 2-4 syll words decline only
  slightly with length
\item
  Prediction based on previous work: Similarly, we do not know of NWR
  research that manipulates the difficulty of the sounds that are
  included in the items, but word naming and other research suggests
  that children achieve higher scores when producing easy and/or
  typologically common sounds than difficult and/or typologically rare
  sounds {[}CITE{]}. Therefore, we expect higher scores for items with
  common sounds than in those with rare sounds.
\item
  Prediction for Yélî made before seeing the data: The Yélî sound
  inventory is very large and compressed, with many similar sounds that
  are acoustic and articulatory neighbors. Therefore, this may
  constitute a pressure for children to have finer auditory skills (and
  perhaps more precise articulations) than children speaking languages
  with a simpler inventory. As a result, differences between easier and
  harder items may be smaller in this work than in other research.
\item
  it turns out we do see
\end{enumerate}

\begin{itemize}
\tightlist
\item
  How frequent are errors that result in real words? Is that a function
  of item complexity?
\item
  Is individual variation explainable by child age, sex, birth order,
  monolingual status, and/or parental education?
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\item
  Children's scores increases with child age.
\item
  General prediction: Non-monolingual Yélî children have lower scores
  than monolingual ones when tested on the society-dominant language (we
  did not test any non-dominant language) 
\item
  Local prediction: Anecdotally Yélî children grow up in close-knitted
  communities and thus may receive significant portions of their
  language input from people not in their nuclear family (or at least
  from people other than their mothers, who tend to be the non-native
  speakers). If so, the difference between monolinguals and not
  monolinguals may be smaller than that found in other work . That said,
  one recent study on the same population shows that most child-directed
  input in the first 2 years does come from the mother , so in so far as
  this input has a crucial formational role, then there may still be a
  score difference between these two groups.
\item
  We don't see a difference
\item
  General prediction: previous NWR evidence on this is mixed, but
  general findings on language development suggest that children whose
  mothers are more educated score higher in standardized language tests
  (eg ppvt) than children whose mothers are less educated.
\item
  Prediction made before seeing data: In the Rossel community, formal
  education plays an extremely minor role in ensuring individual's
  success, is not a good index of relative socio-economic status, and
  furthermore there is only a narrow range of variation in maternal
  educational attainment. This may lead to no or only very small
  advantages for children whose mothers are more educated, provided that
  the causal chain between maternal education and child language is via
  SES more broadly. However, if education directly boosts maternal
  verbal skills and the incidence of verbal behavior (as suggested by
  CITE), then we should still see a difference along this factor.
\item
  We did not find this
\item
  General prediction: To our knowledge, there is no previous NWR work on
  this, but other research suggests that first-born children score
  higher in standardized language tests later-born children.
\item
  Prediction before seeing data: One main causal path between birth
  order and language development is via parental input (CITE). Given our
  arguments above for how mothers may not be as important among Rossel
  people than in other places, then the difference in scores between
  first borns and later borns may be small
\item
  we did not find a sig effect, but this is a small effect (eg d=.2 in
  Havron et al)
\end{enumerate}

>>>>

Next, we investigated patterns of deletion and substitution. Deletions
were relatively rare: there were only 7 instances of deleted vowels
(\textasciitilde{}0.23\% of all vowel targets), and 3 instances of
deleted consonants (\textasciitilde{}0.12\% of all consonant targets).
Substitutions were somewhat more frequent, but still relatively rare.
Nasal vowels were produced as oral vowels and vice versa (34 oral
targets as nasal (\textasciitilde{}0.99\% of targets) and 30 nasal
targets produced as oral (\textasciitilde{}2.37\% of targets)). Both
oral and nasal vowels were also occasionally subsituted by vowels of a
different quality that preserved the target oral/nature airflow (110
oral vowels (\textasciitilde{}1.86\% of targets and 13 nasal vowels
(\textasciitilde{}2.06\% of targets). Among consonants, complex
consonants were occasionally substituted by simpler ones (specifically
{[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and
{[}lβʲ{]} as {[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]},
{[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and
{[}tʃ{]}), 115 times, in approximately 1.96\% of all complex consonant
targets. Interestingly, simple consonants were sometimes also
substituted by complex ones; this happened 2 times, in approximately
1.96\% of all simple consonant targets. Finally, simple consonants were
mispronounced as other simple consonants (68 instances,
\textasciitilde{}0.38\% of all simple consonant targets) and complex
consonants as other complex consonants (33 instances,
\textasciitilde{}1.02\% of all complex consonant targets). Note that the
substitutions included phones that are not native to Yélî Dnye but do
occur in English (e.g., {[}tʃ{]}); as these data come from careful
transcriptions by a native Yélî Dnye speaker who is very fluent in
English, we take these segmental substitutions faithfully as an
indication that several of our participants have mastered production of
some English phones, possibly produced within whole English word forms
(see below).

@@

We used non-word repetition to investigate phonological development in a
language with a large phonological inventory (including some
typologically rare segments). While the study, in itself, advances NWR
research by demonstrating the successful application of this method in a
rural, non-industrialized population (see also Cristia et al., 2020),
the results also inform the larger body of NWR work regarding the
influence of stimulus length, segmental frequency, age, and
developmental context, where previous findings have often been mixed.

In our sample of 40 children between ages 3 and 10, the average
item-level score was 64\%, falling well within the expected range of NWR
applications in previous studies on a variety of linguistic populations
(see Figure \ref{fig:prevlit-fig}). Segment-level analyses revealed that
repetitions recovered an average of 78\% of the segments in the original
stimulus, indicating that inexact repetitions often constituted
\emph{mostly} accurate reroductions of the target. Complementary data
from segment-based normalized Levenshtein distance reflect this same
pattern, with an average of 22\% of the stimulus segments getting
substituted, inserted, or deleted upon repetition. Paired with our
thorough training protocol, we take these NWR scores as indicating that
(a) our adaptations to NWR for this context were successful, even given
a number of non-standard changes to the training phase and to the design
of the stimuli and (b) Yélî children show comparable performance to
others tested on a similar task, despite the many linguistic, cultural,
and socioeconomic differences between this and previously tested
populations. Given successful (and comparable) execution of this task,
we can dive deeper into effects of stimulus length, segmental frequency,
age, and developmental context.

\subsubsection{Item complexity}\label{item-complexity}

We investigated the effect of item complexity on NWR scores by varying
both the number of syllables in the item and its average segmental
frequency. Based on previous work, we had predicted that children would
have higher NWR scores for shorter items. That said, previous work has
shown both very small (Piazzalunga et al., 2019) and very large (Cristia
et al., 2020; Jaber-Awida, 2018) effects of stimulus length and,
further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic
words are nearly equally frequent in the current language, with
trisyllabic and longer words making up a non-trivial 10\% of the
remaining words. Compare this to, for example, English, which is
substantially more skewed toward monosyllabic word forms **Alex I'm
going off your note here (``Prediction for Yélî made before seeing the
data: The length distribution in Yélî words is more balanced than that
in English, and thus the score decline for poly- versus mono-syllables
may be less pronounced than that for English.''``). I don't have a
reference for this, can you please finish the thought or nix this
bit?**. Setting aside our monosyllabic stimuli, which all contained
typologically infrequent segments, we can examine effects of item length
among the remaining stimuli, which range between 2 and 4 syllables long.
While indeed NWR scores were overall lower for longer items (e.g., see
Figure \ref{fig:prevlit-fig}), the effect of item length was not
significant in a statistical model that additionally accounted for age
and random effects of item and participant. In light of mixed prior
results of item length, we propose two possible (and non-mutually
exclusive) explanations for this minimal impact of item length. First,
further extensions of this type of analysis in more populations may
reveal that, in general (and cross-linguistically), item length effects
are variable between languages, potentially reflecting the distribution
of word lengths in the ambient language and other (morpho-)phonological
tendencies in the lexicon. Second, above and beyond these
language-specific effects, the general impact of item length on NWR
score may be relatively small, as shown in Piazzalunga et al.'s (2019)
study on Italian and as borne out in the current dataset once
controlling for other factors.

Our monosyllabic items included typologically rare segments so that we
could test whether lower average segmental frequency is associated with
lower NWR scores. Typologically common sounds are associated with higher
performance on a handful of other tasks (\textbf{REFS -- M2A: Alex, I
added this based on your note, where it sounded like you had some
particular studies in mind?}) though to our knowledge this has not yet
been tested with non-word repetition. Regarding Yélî Dnye in particular,
the phonemic inventory is both large and acoustically packed, in
addition to containing several typologically infrequent (or unique)
contrasts. We therefore expected to see that, while NWR scores would be
lower for stimuli with lower average frequency, this effect would be
relatively weak because the ambient language puts pressure on Yélî
children to distinguish (perceptually and articulatorily) fine-grained
phonetic differences in order to successfully communicate with others.
Indeed we found a robust effect of average segmental frequency on NWR
performance: Even accounting for age and random effects of item and
participant, we see that target words with more frequent segments were
repeated correctly more often. This effect is large, with a magnitude
more than twice the size of the effect of participant age. This
significant effect remains even once also accounting for the frequencies
of these segments in Yélî Dnye child-directed speech, which are
correlated with their typological frequencies. In sum, typological
frequency effects, which have been found in other measurements of
phonological processing. appear to strongly affect NWR performance, and
do not appear mitigated by language-specific pressure to make
finer-grained differences earlier in development.

With respect to the types of errors in repetition made, we did not see
clear patterns to further guide our discussion: base rates of deletion
and substitution were fairly low and the relative distribution of errors
over, e.g., nasal vs.~oral vowels and simple vs.~complex consonants,
revealed no remarkable bias in error types. That said, the lack of a
difference could be due to relative imbalance across our stimuli in the
use of these phonemic features (e.g., we included many more more oral
than nasal targets) and future work should investigate such sources of
error bias more systemtically. Some portion of the errors were
introduced when the participant produced a real word (in Yélî Dnye or
English) in response to the stimulus. Real-word repetitions here made up
two thirds of errorful repetitions---this is quite high compared to past
work (e.g., {\textbf{???}}), but it is unclear what caused this pattern
in the current study: Castro and colleagues' ({\textbf{???}}) study
focused on adults rather than children, the task was administered by a
team including a foreign, English-speaking researcher, and the
particularities of the Yélî Dnye phonological inventory result in many
true-word phonetic neighbors. Follow-up work exploring this type of
error in children from other populations in addition to further work on
Yélî children will clarify this effect.

\subsubsection{Individual differences}\label{individual-differences}

\begin{itemize}
\tightlist
\item
  Is individual variation explainable by child age, sex, birth order,
  monolingual status, and/or parental education?
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\item
  Children's scores increases with child age.
\item
  General prediction: previous NWR evidence on this is mixed, but
  general findings on language development suggest that children whose
  mothers are more educated score higher in standardized language tests
  (eg ppvt) than children whose mothers are less educated.
\item
  Prediction made before seeing data: In the Rossel community, formal
  education plays an extremely minor role in ensuring individual's
  success, is not a good index of relative socio-economic status, and
  furthermore there is only a narrow range of variation in maternal
  educational attainment. This may lead to no or only very small
  advantages for children whose mothers are more educated, provided that
  the causal chain between maternal education and child language is via
  SES more broadly. However, if education directly boosts maternal
  verbal skills and the incidence of verbal behavior (as suggested by
  CITE), then we should still see a difference along this factor.
\item
  We did not find this
\item
  General prediction: To our knowledge, there is no previous NWR work on
  this, but other research suggests that first-born children score
  higher in standardized language tests later-born children.
\item
  Prediction before seeing data: One main causal path between birth
  order and language development is via parental input (CITE). Given our
  arguments above for how mothers may not be as important among Rossel
  people than in other places, then the difference in scores between
  first borns and later borns may be small
\item
  we did not find a sig effect, but this is a small effect (eg d=.2 in
  Havron et al)
\end{enumerate}
